---
title: "Lab05 Extra Credit"
author: "Siddharth K Bailkeri"
format: pdf
editor: visual
---

## Section 7:

Reading in the gapminder files

```{r}
# Downloading package
library(gapminder)
```

Filtering to only rows from the year 2007

```{r}
library(dplyr)

gapminder_2007 <- gapminder %>% filter(year==2007)
```

Creating the First Scatterplot

```{r}
library(ggplot2)
ggplot(gapminder_2007) +
  aes(gdpPercap, lifeExp) +
  geom_point(alpha = 0.5)
```

Life expectancy maps to the y-axis since it is the dependent variable.

### Adding more variables to aes()

```{r}
ggplot(gapminder_2007) +
  aes(gdpPercap, lifeExp, color = continent, size = pop) +
  geom_point(alpha = 0.5)
```

Changing the colouring to match a continuous variable and ncreasing the alpha so we can see indivdual points.

```{r}
ggplot(gapminder_2007) +
  aes(gdpPercap, lifeExp, color = pop) +
  geom_point(alpha = 0.8)
```

### Adjusting point size

```{r}
ggplot(gapminder_2007) +
  aes(gdpPercap, lifeExp, size = pop) +
  geom_point(alpha = 0.5)
```

Changing the point size:

```{r}
ggplot(gapminder_2007) +
  aes(gdpPercap, lifeExp, size = pop) +
  geom_point(alpha = 0.5) +
  scale_size_area(max_size = 10)
```

### Adapting the code for the year 1957

```{r}
gapminder_1957 <- gapminder %>% filter(year == 1957)
ggplot(gapminder_1957) +
  aes(gdpPercap, lifeExp, color = continent, size = pop) +
  geom_point(alpha = 0.7) +
  scale_size_area(max_size = 15)
```

Adapting the code so the graphs wrap

```{r}
gapminder_1957_2007 <- gapminder %>% filter(year == 1957 | year == 2007)
ggplot(gapminder_1957_2007) +
  aes(gdpPercap, lifeExp, color = continent, size = pop) +
  geom_point(alpha = 0.7) +
  scale_size_area(max_size = 15) +
  facet_wrap(~year)
```

The pipe operator followed by the filter() function allows us to filter the dataset by a specific criteria. Here we use \| to add a second criteria to create a dataset with data from both 2007 and 1957.

## Section 8: Bar Charts

Filtering the data to create top 5 by population

```{r}
gapminder_top5 <- gapminder %>%
  filter(year == 2007) %>%
  arrange(desc(pop)) %>%
  top_n(5, pop)

gapminder_top5
```

### Creating the bar chart

```{r}
ggplot(gapminder_top5) +
  aes(x = country, y = pop) +
  geom_col()
```

### Bar chart of life expectancy for five biggest countries by population

```{r}
ggplot(gapminder_top5) +
  aes(x = country, y = lifeExp) +
  geom_col()
```

### Filling bar charts

```{r}
ggplot(gapminder_top5) +
  aes(x = country, y = pop, fill = continent) +
  geom_col()
```

Using a numeric variable instead:

```{r}
ggplot(gapminder_top5) +
  aes(x = country, y = pop, fill = lifeExp) +
  geom_col()
```

**Q. Plot population size by country. Create a bar chart showing the population (in millions) of the five biggest countries by population in 2007.**

```{r}
ggplot(gapminder_top5) +
  aes(x = reorder(country, -pop), y = pop, fill = country) +
  geom_col(col = "black")
```

### Flipping Bar Charts

```{r}
head(USArrests)

USArrests$State <- rownames(USArrests)
ggplot(USArrests) +
  aes(x = reorder(State, Murder), Murder) +
  geom_col() +
  coord_flip()
```

Making it less clunky

```{r}
ggplot(USArrests) +
  aes(x = reorder(State, Murder), Murder) +
  geom_point() +
  geom_segment(aes(x = State, 
                   xend = State,
                   y = 0,
                   yend = Murder),
                   color = "skyblue") +
  coord_flip() +
  labs(title = "Number of Murder Arrests by State", 
       x = "State", 
       y = "Number of Murder Arrests")
```

## Section 9: Animations

```{r}
#library(gganimate)
#library(gapminder)
#setting up the plot
#ggplot(gapminder, aes(gdpPercap, lifeExp, size = pop, #colour = country)) +
  #geom_point(alpha = 0.7, show.legend = FALSE) +
  #scale_colour_manual(values = country_colors) +
  #scale_size(range = c(2, 12)) +
  #scale_x_log10() +
  # Facet by continent
  #facet_wrap(~continent) +
  # animations
  #labs(title = 'Year: {frame_time}', x = 'GDP per capita', y = 'life expectancy') +
  #transition_time(year) +
  #shadow_wake(wake_length = 0.1, alpha = FALSE)
```

This code was inserted as comments since the animated gif is unable to integrate into the pdf. When actually run in RStudio, it ran without error and the output was a gif of life expectancy against GDP per capita animated by year.

## Section 10: Combining Plots

```{r}
library(patchwork)
p1 <- ggplot(mtcars) + geom_point(aes(mpg, disp))
p2 <- ggplot(mtcars) + geom_boxplot(aes(gear, disp, group = gear))
p3 <- ggplot(mtcars) + geom_smooth(aes(disp, qsec))
p4 <- ggplot(mtcars) + geom_bar(aes(carb))

( p1 | p2 | p3 ) / 
  p4
```
