---
title: "Lab09"
author: "Siddharth Bailkeri"
format: pdf
editor: visual
---

## 2. Importing Candy Data

```{r}
url <- "https://raw.githubusercontent.com/fivethirtyeight/data/master/candy-power-ranking/candy-data.csv"

candy <- read.csv(url, row.names = 1)
head(candy)
```

> **Q1. How many different candy types are in this dataset?**

```{r}
nrow(candy)
```

> There are 85 different candy types in the dataset

> **Q2. How many fruity candy types are in the dataset?**

```{r}
sum(candy$fruity == 1)
```

> Ther are 38 fruity types of candy in the candy dataset.

### 2.2 Favourite Candy

```{r}
candy["Twix", ]$winpercent

library(dplyr)
candy |> filter(row.names(candy) == "Twix") |>
  select(winpercent)
```

> **Q3. What is your favorite candy (other than Twix) in the dataset and what is it’s winpercent value?**

```{r}
candy["Snickers", ]$winpercent
```

> **Q4. What is the winpercent value for “Kit Kat”?**

```{r}
candy["Kit Kat", ]$winpercent
```

> **Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?**

```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent
```

### Using skimr::skim()

```{r}
library(skimr)
skim(candy)
```

> **Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?** The winpercent column is on a different scale than the rest .

> **Q7. What do you think a zero and one represent for the candy\$chocolate column?** The zero indicates there are no missing values, i.e. there are o blanks or NAs. The 1 probably indicates that the rate adds up to one, which makes sense because it is represented as a fraction.

## 3. Exploratory Analysis

> **Q8. Plot a histogram of winpercent values using both base R an ggplot2.**

Using base R

```{r}
hist(candy$winpercent)
```

Using ggplot2:

```{r}
library(ggplot2)

ggplot(candy) +
  aes(x = winpercent) +
  geom_histogram(binwidth = 7) +
  theme_bw()
```

> **Q9. Is the distribution of winpercent values symmetrical?**
>
> No, the distribution of winpercent values is not symmetrical.

> **Q10. Is the center of the distribution above or below 50%?**
>
> The center of the distribution is below 50%, around \~40%.

> **Q11. On average is chocolate candy higher or lower ranked than fruit candy?**

```{r}

fruity <- candy$winpercent[as.logical(candy$fruity)]
chocolate <- candy$winpercent[as.logical(candy$chocolate)]

ggplot(candy) +
  aes(x = winpercent, fill = factor(chocolate)) +
  geom_histogram() +
  theme_bw()
```

> Yes, on average, chocolate candy is ranked higher than fruit candy.

> **Q12. Is this difference statistically significant?**

```{r}
fruity <- candy$winpercent[as.logical(candy$fruity)]
chocolate <- candy$winpercent[as.logical(candy$chocolate)]

t.test(chocolate, fruity, var.equal = FALSE, alternative = "greater")
#here, var.equal = FALSE tells R that the two data sets do not have equal variance. 
#alternative = "greater" tells R that we want to test if chocolate 
#(the first argument) is greater than fruity (the second arument)
```

> Yes, since we obtained a p-value of 1.436e-08 from a one-tailed two sample t-test, we can conclude that the difference between chocolate candies and fruity candies is statistically significant, with chocolate being rated higher (hence, the one-tailed test).

## 4. Overall Candy Rankings

> **Q13. What are the five least liked candy types in this set?**

```{r}
head(candy[order(candy$winpercent),], n=5)
```

Using dplyr

```{r}
candy |> arrange(winpercent) |> head(5)
```

> Nik l Nip, Boston Baked Beans, Chiclets, Super Bubble, and Jawbusters are the least liked candies in this dataset.

> **Q14. What are the top 5 all time favorite candy types out of this set?**

```{r}
candy |> arrange(desc(winpercent)) |> head(5)
```

> Reese's peanut butter cups, Reese's miniatures, Twix, Kit Kat, and Snickers are the most liked candies of this dataset.

> **Q15. Make a first barplot of candy ranking based on winpercent values.**

```{r}
ggplot(candy) +
  aes(winpercent, rownames(candy)) +
  geom_col() +
  theme_bw() +
  labs(x = "winpercent", y = "") +
  theme(
    axis.text.y = element_text(size = 4)
  )
```

> **Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?**

```{r}
ggplot(candy) +
  aes(winpercent, reorder(rownames(candy), winpercent)) +
  geom_col() +
  theme_bw() +
  labs(x = "winpercent", y = "") +
  theme(
    axis.text.y = element_text(size = 4)
  )
```

### 4.0.1 Time to add some useful color

```{r}
my_colors <- rep("black", nrow(candy))
my_colors[as.logical(candy$chocolate)] = "chocolate"
my_colors[as.logical(candy$bar)] = "brown"
my_colors[as.logical(candy$fruity)] = "pink"

ggplot(candy) +
  aes(x = winpercent, y = reorder(rownames(candy), winpercent)) +
  geom_col(fill = my_colors) +
  theme_bw() +
  labs(x = "winpercent", y = "") +
  theme(
    axis.text.y = element_text(size = 4)
  )
  

```

> **Q17. What is the worst ranked chocolate candy?**
>
> Sixlets are the worst ranked chocolate candy

> **Q18. What is the best ranked fruity candy?**
>
> Starburst is the best ranked fruity candy

## 5 Taking a look at pricepercent

using ggrepel

```{r}
library(ggrepel)

ggplot(candy)+
  aes(x = winpercent, y = pricepercent, label = rownames(candy)) +
  geom_point(col = my_colors) + 
  geom_text_repel(col = my_colors, size = 3.3, max.overlaps = 5)
```

> **Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?**

> From the graph, reese's miniatures

> **Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?**

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head(candy[ord, c(11,12)], n = 5)
```

> Nik l nip is the most expensive and least popular candy in this data set.

## 6 Exploring the correlation structure

```{r}
library(corrplot)

cij <- cor(candy)
corrplot(cij)
```

> **Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?**

> Chocolate and fruity are anti-correlated

> **Q23. Similarly, what two variables are most positively correlated?**

> chocolate and winpercent were most positively correlated, as were chocolate and bar.

## 7. Principal Component Analysis

```{r}
pca <- prcomp(candy, scale = TRUE)
summary(pca)
```

Plotting the PCA rotations

```{r}
plot(pca$x[ , 1:2])
plot(pca$x[, 1:2], col = my_colors, pch = 16)
```

using ggplot to plot the PCA data

```{r}
mydata <- cbind(candy, pca$x[,1:3])

p <- ggplot(mydata) +
  aes(x = PC1, y = PC2,
      size = winpercent/100,
      text = rownames(mydata), 
      label = rownames(mydata)) +
  geom_point(col = my_colors)
p
```

Using ggrepel

```{r}
p + geom_text_repel(size=3.3, col=my_colors, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), 
       fruity (pink), other (black)",
       caption="Data from 538")
```

Using plotly

```{r}
library(plotly)

ggplotly(p)
```

Generating the loading plot

```{r}
ggplot(pca$rotation) +
  aes(PC1, reorder(rownames(pca$rotation), PC1)) +
  geom_col(fill = "coral1") +
  labs(y = "", x = "PC1") +
  theme_bw()
```

> **Q24. Complete the code to generate the loadings plot above. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you? Where did you see this relationship highlighted previously?**

> The fruity, pluribus, and hard variables are picked up strongly by PC1 in the positive direction. These make sense looking at the plot of PC2 against PC1 since the fruity candies are very high along the x - axis, and so the loading plot for PC1 would show these as highly positive values. We saw this relationsuop highlighted in our correlation plot, so we know that fruity, hard, and pluribus are positively correlated with each other so these also make sense.

## 8. Summary

> **Q25. Based on your exploratory analysis, correlation findings, and PCA results, what combination of characteristics appears to make a “winning” candy? How do these different analyses (visualization, correlation, PCA) support or complement each other in reaching this conclusion?**
> 
> Being chocolatey and being a bar correlated very positively with a high win percent. THe visualization and correlation supported each other very strongly in this, as did the PCA. The correlation plot showed numerically what correlated with a high win percent and the visulaization helped differentiate by which type was a winning candy. The PCA also reinforced the data we obtained from the correlation plot and the data visualization. 
