---
title: "Lab08"
format: pdf
author: "Siddharth Bailkeri"
---

## 1. Preparing the data

```{r}
wisc.df <- read.csv("WisconsinCancer.csv", row.names = 1)
head(wisc.df)
```

Removing the Diagnosis from the data frame and creating a new vector:

```{r}
wisc.data <- wisc.df[ , -1]
diagnosis <- factor(wisc.df$diagnosis)
```

## 2. Exploratory Data Analysis:

```{r}
#finding number of observations
dim(wisc.data)

#finding number of malignant diagnoses
table(diagnosis)

#using grep() to return which data in wisc.data has "_mean" in the column name
#and then counting that up using the length() function
length(grep("_mean", colnames(wisc.data)))

```

> **Q1: How many observations are in this dataset?**

> There are 569 observations in the dataset

> **Q2: How many of the observations have a malignant diagnosis?** There are 212 observations with a malignant diagnosis

> **Q3: How many variables/features in the data are suffixed with "\_mean" ?**

> 10 variables are suffixed with "\_mean" in the data.

## 3. PCA

Checking if the data need to be scaled before PCA:

```{r}
# to find the means:
summary(colMeans(wisc.data))

#finding standard deviation:
summary(apply(wisc.data, 2, sd))
```

Seeing that the ranges of the means and s.d. are so wide, the data needs to be scaled.

### Performing the PCA

```{r}
wisc.pr <- prcomp(wisc.data, scale = T)
summary(wisc.pr)$importance
```

> **Q4. From your results, what proportion of the original variance is captured by the first principal component (PC1)?**

> From my results, 44.272% of variance is captured byt he frist principal component.

```{r}
pca.var <- wisc.pr$sdev^2
pca.var.per <- round(pca.var/sum(pca.var)*100, 3)
pca.var.per

```

> **Q5. How many principal components (PCs) are required to describe at least 70% of the original variance in the data?**

> We need 3 PCs to describe at least 70% of the variation in the data.

```{r}
sum(pca.var.per[1:3])
```

> **Q6. How many principal components (PCs) are required to describe at least 90% of the original variance in the data?**

> We need 7 PCs to describe at least 90% of variance in the data.

```{r}
sum(pca.var.per[1:7])
```

### Interpreting PCA results:

Creating a biplot:

```{r}
biplot(wisc.pr)
```

> **Q7. What stands out to you about this plot? Is it easy or difficult to understand? Why?**

> The plot is really messy with a lot of points overlapping. It is very difficult to understand because it's trying to do a lot of things at once in an overlapping manner.

Using ggplot:

```{r}
library(ggplot2)
ggplot(wisc.pr$x) + 
  aes(x = PC1, y = PC2, color = diagnosis) +
  geom_point(size = 0.8, alpha = 0.5) + 
  labs(title = "PC1 vs PC2", x = "PC1", y = "PC2") +
  theme_bw() + 
  theme(
    axis.title = element_text(size = 10)
  )
```

> **Q8. Generate a similar plot for principal components 1 and 3. What do you notice about these plots?**

> 

```{r}
ggplot(wisc.pr$x) + 
  aes(x = PC1, y = PC3, color = diagnosis) +
  geom_point(size = 0.8, alpha = 0.5) + 
  labs(title = "PC1 vs PC3", x = "PC1", y = "PC3") +
  theme_bw() + 
  theme(
    axis.title = element_text(size = 10)
  )
```

### Variance explained

Calculating proportion of variance

```{r}
# variance calculated above to answer questions 4 - 6
head(pca.var)
```

Plotting proportion of variance:

```{r}
#calculating proportion of variance
pve <- pca.var/sum(pca.var)

#plotting this:
plot(c(1,pve), xlab = "Principal Component", 
     ylab = "Proportion of Variance Explained", 
     ylim = c(0, 1), type = "o")
```

Alternate plot:

```{r}
barplot(pve, ylab = "Percent of Variance Explained",
        names.arg=paste0("PC",1:length(pve)), las=2, axes = FALSE)
axis(2, at=pve, labels=round(pve,2)*100 )
```

Using factoextra

```{r}
library(factoextra)
fviz_eig(wisc.pr, addlabels = TRUE)
```

### Communicating PCA results

Generating a loading plot for PC1:

```{r}
ggplot(wisc.pr$rotation) +
  aes(x = PC1, y = reorder(rownames(wisc.pr$rotation), PC1)) +
  geom_col(fill = "orange") +
  labs(title = "PC1 Loading Plot", 
       x = "PC1 Loading Score", 
       y = "") +
  theme_bw() +
  theme(axis.text.y = element_text(size = 9))
```

> **Q9. For the first principal component, what is the component of the loading vector (i.e. wisc.pr\$rotation\[,1\]) for the feature concave.points_mean? This tells us how much this original feature contributes to the first PC. Are there any features with larger contributions than this one?**

> The feature concave.points_mean contributes -0.2608538 to the mean. There are no features with larger contributions than this one, as seen by the plot above and also the ordered list below.

```{r}
head(sort(wisc.pr$rotation[,1], decreasing = FALSE))
```

## 4. Hierarchical Clustering

### Results of hierarchical clustering

Scaling the data:

```{r}
data.scaled <- scale(wisc.data)
data.dist <- dist(data.scaled)
wisc.hclust <- hclust(data.dist, method = "complete")
```

> **Q10. Using the plot() and abline() functions, what is the height at which the clustering model has 4 clusters?**

> At around y = 20, the model has 4 clusters.

```{r}
plot(wisc.hclust)
abline(h = 20, col = "red", lty = 2)
```

### Selecting number of clusters

```{r}
wisc.hclust.clusters <- cutree(wisc.hclust, k = 4)
table(wisc.hclust.clusters)
```

Using table()

```{r}
table(wisc.hclust.clusters, diagnosis)
```

> **Q11. OPTIONAL: Can you find a better cluster vs diagnoses match by cutting into a different number of clusters between 2 and 6? How do you judge the quality of your result in each case?**

> No, i could not find any number fo clusters between 2-6 that works better than k = 4 because anything k \< 4 has almost equal malignant and benign cells in the same cluster and k \> 4 has some groups with almost nothing in them.

```{r}
repeat.clusters <- function(k) {
  table(cutree(wisc.hclust, k = k), diagnosis)}

sapply(c(2,3,5,6), repeat.clusters)
```

```{r}
repeat.method <- function(x){
  plot(hclust(data.dist, method = x))}

sapply(c("single", "complete", "average", "ward.D2"), repeat.method)
```

```{r}
table(cutree(hclust(data.dist, method = "ward.D2"), k = 2), diagnosis)
```

> **Q12. Which method gives your favorite results for the same data.dist dataset? Explain your reasoning.**

> The ward.D2 method gives me my favorite results for the same dataset because when i break the data in 2 clusters after applying this method fro hclust(), the resulting groups correspond very closely to the diagnosis of B or M.

## 5. Combining methods

### Clustering on PCA results

```{r}
wisc.pr.hclust <- hclust(dist(wisc.pr$x[ ,1:7]), method = "ward.D2")
plot(wisc.pr.hclust)
```

Visulaizing accuracy:

```{r}
grps <- cutree(wisc.pr.hclust, k = 2)
table(grps, diagnosis)
```

Plotting using ggplot:

```{r}
ggplot(wisc.pr$x) +
  aes(PC1, PC2) +
  geom_point(color = grps, alpha = 0.7, size = 0.8) +
  theme_bw()
```

### Optional: 3D plots using rgl

```{r}
#library(rgl)
#plot3d(wisc.pr$x, xlab = "PC1", ylab = "PC2", zlab = "PC3", size = 1, col = 
#grps, type = "s", cex = 1.5)
#rglwidget(width = 400, height = 400)
```

> **Q13. How well does the newly created hclust model with two clusters separate out the two “M” and “B” diagnoses?**

> The newly created model differentiates between M and B better since there are fewer clusters with more accurate groupings. In the previous hclust, there were 2 groups with almost no observations but fewer clusters resulted in inaccurate groupings. However, this hclust also shows slightly increased false positive and false negative readouts.

> **Q14. How well do the hierarchical clustering models you created in the previous sections (i.e. without first doing PCA) do in terms of separating the diagnoses? Again, use the table() function to compare the output of each model (wisc.hclust.clusters and wisc.pr.hclust.clusters) with the vector containing the actual diagnoses.**

> The previous hclust models didn't do very well in differentiating the actual diagnoses since there was a larger number of groupings required in hclust than the actual diagnoses. It also included more false positives for malignant cells. 

```{r}
table(wisc.hclust.clusters, diagnosis)
table(grps, diagnosis)
```


## 6. Specificity

> **Q15. OPTIONAL: Which of your analysis procedures resulted in a clustering model with the best specificity? How about sensitivity?**

> 

```{r}
best.model <- table(grps, diagnosis)
sensitivity <- sum(best.model[ ,2])/sum(best.model[1,2], best.model[1,])
specificity <- sum(best.model[ ,1])/sum(best.model[1,1], best.model[2,])
best.model
```

## 7. Prediction:

Using predict() to apply our PCA model to new data:

```{r}
new.data <- read.csv("new_samples.csv")
npc <- predict(wisc.pr, newdata = new.data)
npc
```

Plotting the prediction of new data:

```{r}
plot(wisc.pr$x[ , 1:2], col = grps)
points(npc[ ,1], npc[ ,2], col = "blue", pch = 16, cex = 3)
text(npc[,1], npc[,2], c(1,2), col = "white")
```

> **Q16. Which of these new patients should we prioritize for follow up based on your results?**

> We should prioritize following up with Patient 2 since their cell nuclei measurements make them fall in the malignant category.
